# Multilevel Modeling

## Multilevel data

We have simulated data from 100 classes, with a different number of pupils in each class. The average class size is 20 pupils. On the pupil level, we have two variables. First is the dependent variable 'popularity', measured on a self-rating scale that ranges from 0 (very unpopular) to 10 (very popular). Second is the independent variable 'extraversion', measured on a self-rating scale ranging from 1 to 10. On the class level, we have one explanatory variable 'teacher experience', measured in years ranging from 2 to 25.

## Building the multilevel regression model

We are to build three multilevel regression models in this practical. The three models to be built are as follows:

- Empty model (aka. intercept-only model)
- Model with a level-1 predictor
- Model with a level-2 predictor

For simplicity and illustrative purposes, we only consider random intercepts but not random slopes. On Wednesday, you will build the same three models in HLM software using a different dataset.

## Main questions

### Empty model (aka. intercept-only model)

1. Write Mplus syntax.

```{r, eval=FALSE}
! Annotate what you are doing in this line
TITLE: Empty model (intercept-only model)

DATA:
	! Annotate what you are doing in this line
	file is popular2.dat;

VARIABLE:
	! Annotate what you are doing in this line
	names are class pupil cons extrav sex texp popular popteach zextrav
	zsex ztexp zpopular zpoptch;
	
	! Annotate what you are doing in this line
	usevariables are popular;
	
	! Annotate what you are doing in this line
	cluster is class;

ANALYSIS:
	! Annotate what you are doing in this line
	type is twolevel;
	
	! Annotate what you are doing in this line
	estimator is MLR;

MODEL:
	! Annotate what you are doing in this line
	%within%

	! Annotate what you are doing in this line
	%between%

OUTPUT:
	! Annotate what you are doing in this line
	sampstat cinterval;
```

2. Run the analysis and interpret the results.

3. Can you match the parameter estimates to the notation in the formula below?

$$
\begin{gathered}
\textbf{Level 1} \\ 
y_{ij}=\beta_{0j} + r_{ij} \\ \\
\textbf{Level 2} \\ 
\beta_{0j}=\gamma_{00} + u_{0j}
\end{gathered}
$$

4. Can you compute the ICC?

### Model with a level-1 predictor

1. Write Mplus syntax.

```{r, eval=FALSE}
! Annotate what you are doing in this line
TITLE: Model with a level-1 predictor

DATA:
	! Annotate what you are doing in this line
	file is popular2.dat;

VARIABLE:
	! Annotate what you are doing in this line
	names are class pupil cons extrav sex texp popular popteach zextrav
	zsex ztexp zpopular zpoptch;
	
	! Annotate what you are doing in this line
	usevariables are extrav popular;
	
	! Annotate what you are doing in this line
	cluster is class;
	
	! Annotate what you are doing in this line
	within are extrav;

ANALYSIS:
	! Annotate what you are doing in this line
	type is twolevel;
	
	! Annotate what you are doing in this line
	estimator is MLR;

MODEL:
	! Annotate what you are doing in this line
	%within%
	popular on extrav;

	! Annotate what you are doing in this line
	%between%
	popular;

OUTPUT:
	! Annotate what you are doing in this line
	sampstat cinterval;

SAVEDATA:
	! Annotate what you are doing in this line
	file is fscores.dat;
	
	! Annotate what you are doing in this line
	save is fscores;
```

2. Run the analysis and interpret the results.

3. Can you match the parameter estimates to the notation in the formula below?

$$
\begin{gathered}
\textbf{Level 1} \\
y_{ij}=\beta_{0j} + \color{red}{\beta_{1j}X_{ij}} + r_{ij} \\ \\
\textbf{Level 2} \\
\beta_{0j}=\gamma_{00} + u_{0j} \\
\color{red}{\beta_{1j}=\gamma_{10}}
\end{gathered}
$$

4. Can you compute the ICC?

### Model with a level-2 predictor

1. Write Mplus syntax.

```{r, eval=FALSE}
! Annotate what you are doing in this line
TITLE: Model with a level-2 predictor

DATA:
	! Annotate what you are doing in this line
	file is popular2.dat;

VARIABLE:
	! Annotate what you are doing in this line
	names are class pupil cons extrav sex texp popular popteach zextrav
	zsex ztexp zpopular zpoptch;
	
	! Annotate what you are doing in this line
	usevariables are texp popular;
	
	! Annotate what you are doing in this line
	cluster is class;
	
	! Annotate what you are doing in this line
	between are texp;

ANALYSIS:
	! Annotate what you are doing in this line
	type is twolevel;
	
	! Annotate what you are doing in this line
	estimator is MLR;

MODEL:
	! Annotate what you are doing in this line
	%within%
	popular;

	! Annotate what you are doing in this line
	%between%
	popular on texp;

OUTPUT:
	! Annotate what you are doing in this line
	sampstat cinterval;
```

2. Run the analysis and interpret the results.

3. Can you match the parameter estimates to the notation in the formula below?

$$
\begin{gathered}
\textbf{Level 1} \\
y_{ij}=\beta_{0j} + r_{ij} \\ \\
\textbf{Level 2} \\
\beta_{0j}=\gamma_{00} + \color{red}{\gamma_{01}X_{j}} + u_{0j}
\end{gathered}
$$

4. Can you compute the ICC?

## Bonus questions

1. Can you visualize the results of multilevel modeling? Let's practice with the model with a level-1 predictor that we fitted.

```{r}
# Clean the work space
rm(list=ls()); gc()

# Load required packages
library(MplusAutomation)
library(texreg)
library(ggplot2)

# Extract fscores
out <- readModels("2. Model with a l1 pred.out", recursive = FALSE, 
                  what = "savedata")
fscores <- out$savedata
fscores_classid <- aggregate(fscores[,3:4], list(fscores$CLASS), mean)

# Check the range of the predictor
range(fscores$EXTRAV) # 1 to 10

# Find predicted values for each level of extraversion
pred_popular_classid <- data.frame(fscores_classid[rep(seq(nrow(fscores_classid)),
                                                       each = 10),],
                                   "extraversion" = 1:10)
pred_popular <- pred_popular_classid$B_POPULAR + 0.486 * pred_popular_classid$extraversion
pred_popular_classid <- data.frame(pred_popular_classid, pred_popular)

# Visualization
ggplot(data = pred_popular_classid, aes(x = extraversion, y = pred_popular, group = Group.1)) +
  geom_line(aes(color = Group.1), show.legend = FALSE) +
  labs(x = "Extraversion",
       y = "Popularity") +
  scale_y_continuous(lim = c(0, 10), breaks = seq(0, 10, by = 1)) +
  scale_x_continuous(lim = c(1, 10), breaks = seq(1, 10, by = 1)) +
  theme_bw()
```

2. How can you interpret the plot?

