# Introduction to Mplus

## What is it? Why called Mplus?

Mplus is a statistical modeling program that provides researchers with a flexible tool to analyze data

- Many models: regression, path analysis, factor analysis, SEM, MLM, longitudinal models, mixture model, mediation/moderation

- Many data: cross-sectional, longitudinal, single-/multilevel, observed/latent, incomplete

- Many variables: continuous, dichotomous, categorical, count

- Many estimator: maximum likelihood, weighted least squares, Bayesian

## Syntax-based programming

- Commands and subcommands (https://www.statmodel.com/language.html)

- Examples of commands? (https://www.youtube.com/watch?v=XeRRtdmu23k)

   - We will be 'mostly' using TITLE, DATA, VARIABLE, ANALYSIS, MODEL, OUTPUT commands
   - But we will also be often using DEFINE, SAVEDATA, PLOT, MONTECARLO commands

## Some tips when programming

1. Comments can be added with exclamation marks (!)

2. Commands should end with colon (:), and subcommands should end with semicolon (;)

3. Syntax is not case sensitive

4. Data should consist of numeric values, with no variable names

5. Data and Mplus input file should be in the same directory (like an R working directory)

- Otherwise, be sure to specify the correct directory

## Some tips about model command particularly

1. Start with a path diagram

2. Think of it as specifying model parameters

3. Care to the degrees of freedom (DF)

## Example: Multiple linear regression using maximum likelihood estimation

### Model syntax

```{r, eval=FALSE}
! Title command
TITLE: Predicting album sales using ML multiple regression

! Data command
DATA:
	! When data and input file are in the same working directory
	FILE IS Album Sales.csv; ! Subcommands should end with ;

	! When data and input file are in the different working directory
	! FILE IS c:\desktop\different folder\Album Sales.csv;

! Variable command
VARIABLE:
	! Column names (i.e., ALL variable names)
	NAMES ARE adverts sales airplay attract;
	
	! Variables that will be used in our analysis
	USEVARIABLES ARE adverts sales airplay;
	
! Analysis command
ANALYSIS:
	ESTIMATOR IS ML; ! This is the default

! Model command
MODEL:
	! Let's predict sales using adverts and airplay
	! We regress sales on adverts and airplay
	sales ON adverts airplay;

	! If you do not specify variances of and covariances between predictors
	! degrees of freedom (DF) are not correct
	! Variances of exogenous variable
	adverts airplay;
	! Covariances between exogenous variable
	adverts WITH airplay;

! Output command
OUTPUT:
	TECH1 SAMPSTAT STDYX;
	! TECH1 to understand which parameters are being estimated
	! SAMPSTAT to check sample descriptive statistics
	! STDYX to standardize Y (i.e., DV) and X (i.e., IV)
```

### Part of the output file

```{r, eval=FALSE}
MODEL RESULTS

                                                    Two-Tailed
                    Estimate       S.E.  Est./S.E.    P-Value

 SALES    ON
    ADVERTS            0.087      0.007     12.082      0.000
    AIRPLAY            3.589      0.285     12.608      0.000

 ADVERTS  WITH
    AIRPLAY          604.061    421.412      1.433      0.152

 Means
    ADVERTS          614.412     34.255     17.936      0.000
    AIRPLAY           27.500      0.865     31.777      0.000
```

## Additional materials

1. Official website at https://www.statmodel.com/

2. User's guide and examples at https://www.statmodel.com/ugexcerpts.shtml <br> $\rightarrow$ Highly recommended!

3. Mplus YouTube channel at https://www.youtube.com/c/MplusVideos

4. QuantFish YouTube channel at https://www.youtube.com/c/QuantFish

5. Tutorials by Prof. Rens van de Schoot and his students at https://www.rensvandeschoot.com/tutorials/
